Estimating the replicability of psychology experiments after an initial failure to replicate

### 2024-04-17 manuscript/manuscript.Rmd
more typos

---
title: "Estimating the replicability of psychology experiments after an initial failure to replicate"
author:
  - name: Veronica Boyce
    affiliation: Psych
    footnote:
      - corresp
  - name: Ben Prystawski
    affiliation: Psych
  - name: Adani Abutto
    affiliation: Psych
  - name: Emily Chen
    affiliation: Psych
  - name: Ziwen Chen
    affiliation: GSB
  - name: Howard Chiu
    affiliation: Edu
  - name: Irmak Ergin
    affiliation: Psych
  - name: Anmol Gupta
    affiliation: Psych
  - name: Chuqi Hu
    affiliation: Symsys
  - name: Bendix Kemmann
    affiliation: Philosophy
  - name: Nastasia Klevak
    affiliation: Psych
  - name: Verity Y. Q. Lua
    affiliation: Psych
  - name: Mateus M. Mazzaferro
    affiliation: Edu
  - name: Khaing Mon
    affiliation: Symsys
  - name: Dan Ogunbamowo
    affiliation: Psych
  - name: Alexander Pereira
    affiliation: Philosophy
  - name: Jordan Troutman
    affiliation: CS
  - name: Sarah Tung
    affiliation: Psych
  - name: Raphael Uricher
    affiliation: Psych
  - name: Michael C. Frank
    affiliation: Psych
address:
  - code: Psych
    address: Department of Psychology, Stanford University
  - code: GSB
    address: Graduate School of Business, Stanford University
  - code: Edu
    address: Graduate School of Education, Stanford University
  - code: Symsys
    address: Symbolic Systems Program, Stanford University
  - code: Philosophy
    address: Department of Philosophy, Stanford University
  - code: CS
    address: Department of Computer Science, Stanford University
footnote:
  - code: corresp
    text: "Corresponding author. Email: vboyce@stanford.edu"
bibliography: ["251rescue.bib"] # Replace with one or more of your own bibtex files. Better BibTeX for Zotero is your friend
csl: apa7.csl # Use any CSL
nocite: |
  @krauss2003, @ngo2019, @todd2016, @jara-ettinger2022, @porter2016, @birch2007, @child2018, @chou2016, @craig2014, @gong2019, @haimovitz2016, @hopkins2016, @paxton2012, @payne2008, @schechtman2010, @tarampi2016, @yeshurun2003
link-citations: TRUE
output:
  bookdown::pdf_document2:
    toc: FALSE
    keep_tex: TRUE
    template: generic_article_template.tex
    #md_extensions: "-autolink_bare_uris"
    number_sections: TRUE
    citation_package: default # Can also be "natbib"
lang: en # Main document language in BCP47 format
geometry: "margin=25mm"
papersize: a4
#linestretch: 2 # for double spacing
endfloat: FALSE # Set to TRUE to turn on latex endfloat package to place figures and tables at end of document
# endfloatoption: # See endfloat documentation for more possibilities
numberlines: FALSE
authblk: TRUE # FALSE = author affiliations in footnotes; TRUE = author affiliations in a block below author names
footnotehyper: FALSE # TRUE will give you enhanced table footnote capabilities. Set to FALSE to be able to use French blocks. Needed due to what appears to be a latex bug.
urlcolor: blue
linkcolor: blue
citecolor: blue
graphics: TRUE # Needed to be able to include images
tables: TRUE # Needed to be able to include tables
# fancyhdr:
#   first:
#     #headleft: "REPORT-NO-XXXX"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 0pt
#     #footleft: A left foot
#     footrulewidth: 0pt
#   subsequent:
#     #headleft: "NEXT-PAGE-HEADER-LEFT"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 1pt
#     footrulewidth: 0pt
header-includes:
 - \usepackage{tikz}
 - \usetikzlibrary{positioning,chains}
 - \usepackage{setspace}\singlespacing
 - \renewcommand{\textfraction}{0.00}
 - \renewcommand{\topfraction}{1}
 - \renewcommand{\bottomfraction}{1}
 - \renewcommand{\floatpagefraction}{1}
 - \setcounter{topnumber}{3}
 - \setcounter{bottomnumber}{3}
 - \setcounter{totalnumber}{4}
editor_options: 
  markdown: 
    wrap: sentence
---

@@

For the goals of large-scale replication projects, it does not matter why any particular study failed to replicate, because the aim is to estimate the proportion of replication effects in a literature. In contrast, however, when individual scientists attempt to replicate an effect with the goal of building on it, they do care about why an individual replication failed. In an uncertain literature, scientists might not want to commit resources to studying extensions of an effect without first checking that they can measure that effect themselves. For those individual scientists, it is much more important whether the replication failure is due to a fixable issue with the the replication or an intrinsic feature of the original result.

@@

The rescue of @krauss2003 was successful despite an unsuccessful replication.
This study looked at the influence of guided thinking on whether or not people gave correct justifications (drawn or written) for their answer on the Monty Hall problem.
The original paper reported correct justification from 2/67 (`r round(2/67*100)`%) in the control condition and 13/34 (`r round(13/34*100)`%) in the guided thinking condition.
The first replication struggled to recruit participants who were naive to the problem (an exclusion criterion), and many participants give very short text responses in the provided text box (only textual responses were allowed).
The replication found 0/8 correct justifications in the control and 0/11 in the guided thinking condition.
While we cannot know for sure what caused the non-replication, there were clear problems observable from the small final sample and low-quality responses.
The rescue targeted these issues by adding a pre-screening that ensured participants were not familiar with the Monty Hall problem, switched the name of the problem (to reduce googling for answers), and had participants upload drawings for their justifications.
Collectively, we believe these changes brought the rescue closer to the intent of the original.
The rescue had 1/40 (`r round(1/40*100)`%) correct justifications in the control condition and 6/35 (`r round(6/35*100)`%) in the guided thinking condition.
This effect was significant, though smaller than the effect in the original.

@ngo2019 was a second successful rescue. The original study found a large effect; hence the first replication, powering for 80% power on the reported effect, recruited a small sample of 12 people. This small study failed to find the original effect.
The rescue, powered using 2.5x the original sample [as recommended by @simonsohn2015], recovered a clear effect (albeit a much smaller one).
There are reasons to think that some effect sizes in the literature may be inflated [@openscienceconsortium; @camerer2018], and it is also possible that changes to experiments or switches to online could result in noisier samples and thus smaller effect sizes.
Therefore, replications with smaller samples than the original (even if powered to the original effect size), may not be very diagnostic, and could potentially benefit from a re-replication.

>>>>

---
title: "Estimating the replicability of psychology experiments after an initial failure to replicate"
author:
  - name: Veronica Boyce
    affiliation: Psych
    footnote:
      - corresp
  - name: Ben Prystawski
    affiliation: Psych
  - name: Adani B. Abutto
    affiliation: Psych
  - name: Emily M. Chen
    affiliation: Psych
  - name: Ziwen Chen
    affiliation: GSB
  - name: Howard Chiu
    affiliation: Edu
  - name: Irmak Ergin
    affiliation: Psych
  - name: Anmol Gupta
    affiliation: Psych
  - name: Chuqi Hu
    affiliation: Symsys
  - name: Bendix Kemmann
    affiliation: Philosophy
  - name: Nastasia Klevak
    affiliation: Psych
  - name: Verity Y. Q. Lua
    affiliation: Psych
  - name: Mateus M. Mazzaferro
    affiliation: Edu
  - name: Khaing Mon
    affiliation: Symsys
  - name: Dan Ogunbamowo
    affiliation: Psych
  - name: Alexander Pereira
    affiliation: [Psych, Philosophy]
  - name: Jordan Troutman
    affiliation: CS
  - name: Sarah Tung
    affiliation: Psych
  - name: Raphael Uricher
    affiliation: Psych
  - name: Michael C. Frank
    affiliation: Psych
address:
  - code: Psych
    address: Department of Psychology, Stanford University
  - code: GSB
    address: Graduate School of Business, Stanford University
  - code: Edu
    address: Graduate School of Education, Stanford University
  - code: Symsys
    address: Symbolic Systems Program, Stanford University
  - code: Philosophy
    address: Department of Philosophy, Stanford University
  - code: CS
    address: Department of Computer Science, Stanford University
footnote:
  - code: corresp
    text: "Corresponding author. Email: vboyce@stanford.edu"
bibliography: ["251rescue.bib"] # Replace with one or more of your own bibtex files. Better BibTeX for Zotero is your friend
csl: apa7.csl # Use any CSL
nocite: |
  @krauss2003, @ngo2019, @todd2016, @jara-ettinger2022, @porter2016, @birch2007, @child2018, @chou2016, @craig2014, @gong2019, @haimovitz2016, @hopkins2016, @paxton2012, @payne2008, @schechtman2010, @tarampi2016, @yeshurun2003
link-citations: TRUE
output:
  bookdown::pdf_document2:
    toc: FALSE
    keep_tex: TRUE
    template: generic_article_template.tex
    #md_extensions: "-autolink_bare_uris"
    number_sections: TRUE
    citation_package: default # Can also be "natbib"
lang: en # Main document language in BCP47 format
geometry: "margin=25mm"
papersize: a4
#linestretch: 2 # for double spacing
endfloat: FALSE # Set to TRUE to turn on latex endfloat package to place figures and tables at end of document
# endfloatoption: # See endfloat documentation for more possibilities
numberlines: FALSE
authblk: TRUE # FALSE = author affiliations in footnotes; TRUE = author affiliations in a block below author names
footnotehyper: FALSE # TRUE will give you enhanced table footnote capabilities. Set to FALSE to be able to use French blocks. Needed due to what appears to be a latex bug.
urlcolor: blue
linkcolor: blue
citecolor: blue
graphics: TRUE # Needed to be able to include images
tables: TRUE # Needed to be able to include tables
# fancyhdr:
#   first:
#     #headleft: "REPORT-NO-XXXX"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 0pt
#     #footleft: A left foot
#     footrulewidth: 0pt
#   subsequent:
#     #headleft: "NEXT-PAGE-HEADER-LEFT"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 1pt
#     footrulewidth: 0pt
header-includes:
 - \usepackage{tikz}
 - \usetikzlibrary{positioning,chains}
 - \usepackage{setspace}\singlespacing
 - \renewcommand{\textfraction}{0.00}
 - \renewcommand{\topfraction}{1}
 - \renewcommand{\bottomfraction}{1}
 - \renewcommand{\floatpagefraction}{1}
 - \setcounter{topnumber}{3}
 - \setcounter{bottomnumber}{3}
 - \setcounter{totalnumber}{4}
editor_options: 
  markdown: 
    wrap: sentence
---

@@

For the goals of large-scale replication projects, it does not matter why any particular study failed to replicate, because the aim is to estimate the proportion of replication effects in a literature. In contrast, however, when individual scientists attempt to replicate an effect with the goal of building on it, they do care about why an individual replication failed. In an uncertain literature, scientists might not want to commit resources to studying extensions of an effect without first checking that they can measure that effect themselves. For those individual scientists, it is much more important whether the replication failure is due to a fixable issue with  the replication or an intrinsic feature of the original result.

@@

The rescue of @krauss2003 was successful despite an unsuccessful replication.
This study looked at the influence of guided thinking on whether or not people gave correct justifications (drawn or written) for their answer on the Monty Hall problem.
The original paper reported correct justification from 2/67 (`r round(2/67*100)`%) in the control condition and 13/34 (`r round(13/34*100)`%) in the guided thinking condition.
The first replication struggled to recruit participants who were naive to the problem (an exclusion criterion), and many participants gave very short text responses in the provided text box (only textual responses were allowed).
The replication found 0/8 correct justifications in the control and 0/11 in the guided thinking condition.
While we cannot know for sure what caused the non-replication, there were clear problems observable from the small final sample and low-quality responses.
The rescue targeted these issues by adding a pre-screening that ensured participants were not familiar with the Monty Hall problem, reskinned the problem and changed its name (to reduce googling for answers), and had participants upload drawings for their justifications.
Collectively, we believe these changes brought the rescue closer to the intent of the original.
The rescue had 1/40 (`r round(1/40*100)`%) correct justifications in the control condition and 6/35 (`r round(6/35*100)`%) in the guided thinking condition.
This effect was significant, though smaller than the effect in the original.

@ngo2019 was a second successful rescue. The original study found a large effect; hence the first replication, powering for 80% power on the reported effect, recruited a small sample of 12 people. This small study failed to find the original effect.
The rescue, powered using 2.5x the original sample [as recommended by @simonsohn2015], recovered a clear effect (albeit a much smaller one).
There are reasons to think that some effect sizes in the literature may be inflated [@openscienceconsortium; @camerer2018], and it is also possible that changes to experiments or switches to online could result in noisier samples and thus smaller effect sizes.
Therefore, replications with smaller samples than the original (even if powered to the original effect size), may not be very diagnostic and could potentially benefit from a re-replication.
